Computational methods in landscape ecology

### 2024-09-13 Manuscript.qmd
adds some comments, changes, etc

papersize: a4paper
fontsize: 10pt
geometry: margin=25mm
indent: 2m
linestretch: 1.5
number-sections: true

@@

{{< include Sections/01-introduction.qmd >}}
{{< include Sections/02-spatial-patterns.qmd >}}
{{< include Sections/03-connectivity.qmd >}}
{{< include Sections/04-sampling.qmd >}}
{{< include Sections/05-simulations.qmd >}}
{{< include Sections/06-spatial-planning.qmd >}}
{{< include Sections/07-genetics.qmd >}}
{{< include Sections/08-discussion.qmd >}}

>>>>

papersize: a4paper
fontsize: 12pt
geometry: margin=25mm
indent: 2m
linestretch: 1.5
number-sections: true

@@

{{< include Sections/01-introduction.qmd >}}
{{< include Sections/02-spatial-patterns.qmd >}}
{{< include Sections/03-connectivity.qmd >}}
{{< include Sections/04-sampling.qmd >}}
{{< include Sections/05-simulations.qmd >}}
<!-- JN: I would suggest switching the order of spatial planning and genetics -->
{{< include Sections/06-spatial-planning.qmd >}}
{{< include Sections/07-genetics.qmd >}}
{{< include Sections/08-discussion.qmd >}}

### 2024-09-13 Sections/03-connectivity.qmd
adds some comments, changes, etc

Connectivity describes how movement among resources patches is facilitated or hindered by landscape features such as types, amounts, and arrangement of LULC [@Taylor1993ConnectivityVital; @Taylor2006LandscapeConnectivity].
While structural connectivity focuses on physical and spatial relationships among resource patches, functional connectivity also includes the movement or flow of individuals and populations through the landscape [@Taylor2006LandscapeConnectivity].
Connectivity underpins ecosystem functionality and maintains biodiversity and populations, and plays a large role in many conservation actions [@CorreaAyram2016HabitatConnectivity; @Hilty2020GuidelinesConserving].
Thus, in the past multiple approaches how to measure connectivity were developed [@Simpkins2018AssessingPerformance; @Keeley2021ConnectivityMetrics].

@@

## Landscape metrics

There are several landscape metrics that are traditionally used to measure landscape connectivity.
For example, the habitat amount within a buffer zone of focal patches, the nearest-neighbor distance, or the connectance index [@Moilanen2002SimpleConnectivity; @Prugh2009EvaluationPatch; @Spanowicz2019MeasuringLandscape].
However, these metrics are frequently criticized for being to simple and especially the link to functional connectivity is often ambiguous. 
Thus, more functionality needs to be included using additional ecological meaningful metrics, such as the core area [e.g., @Kupfer2012LandscapeEcology].
More recently there is also an increasing focus in within-patch connectivity using metrics such as the effective mesh size in addition to the usual measure between-patch connectivity [@Spanowicz2019MeasuringLandscape].

@@

Resistance surfaces are one of the most popular approaches to model connectivity and developed from simple isolation-by-distance approaches. 
The approach uses environmental layers and biodiversity data or expert opinions to estimate the cost and of movement through landscapes [@Zeller2012EstimatingLandscape].
The latest developments of resistance surfaces have recently been reviewed [@Dutta2022OverviewComputational].
The resistance surface can be used to model movement based on different approaches such as leas-cost paths [@Adriaensen2003ApplicationLeastcost; @Etherington2016LeastcostModelling] or circuit theory [@McRae2008UsingCircuit].
However, resistance surfaces are not free of critique including missing spatial-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape].
Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifterExtended; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], using genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or computational improvements [@Leonard2017GlowSoftware].

>>>>

Connectivity describes how movement among resources patches is facilitated or hindered by landscape features such as types, amounts, and arrangement of LULC [@Taylor1993ConnectivityVital; @Taylor2006LandscapeConnectivity].
While structural connectivity focuses on physical and spatial relationships among resource patches, functional connectivity also includes the movement or flow of individuals and populations through the landscape [@Taylor2006LandscapeConnectivity].
Connectivity underpins ecosystem functionality and maintains biodiversity and populations, and plays a large role in many conservation actions [@CorreaAyram2016HabitatConnectivity; @Hilty2020GuidelinesConserving].
Thus, in the past multiple approaches how to computationally measure connectivity were developed [@Simpkins2018AssessingPerformance; @Keeley2021ConnectivityMetrics].

@@

## Landscape metrics related to connectivity

There are several landscape metrics that are traditionally used to measure landscape connectivity.
For example, the habitat amount within a buffer zone of focal patches, the nearest-neighbor distance, or the connectance index [@Moilanen2002SimpleConnectivity; @Prugh2009EvaluationPatch; @Spanowicz2019MeasuringLandscape].
However, these metrics are frequently criticized for being to simple and especially the link to functional connectivity is often ambiguous. 
Thus, more functionality needs to be included using additional ecological meaningful metrics, such as the core area [e.g., @Kupfer2012LandscapeEcology].
More recently there is also an increasing focus in within-patch connectivity using metrics such as the effective mesh size in addition to the usual measure between-patch connectivity [@Spanowicz2019MeasuringLandscape].
<!--JN: software??-->

@@

Resistance surfaces are one of the most popular approaches to model connectivity and developed from simple isolation-by-distance approaches. 
This concept uses environmental layers and biodiversity data or expert opinions to estimate the cost and of movement through landscapes [@Zeller2012EstimatingLandscape].
The latest developments of resistance surfaces have recently been reviewed by @Dutta2022OverviewComputational.

The resistance surface can be used to model movement based on different approaches such as leas-cost paths [@Adriaensen2003ApplicationLeastcost; @Etherington2016LeastcostModelling] or circuit theory [@McRae2008UsingCircuit].
However, resistance surfaces are not free of critique including missing spatial-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape].
Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifterExtended; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], using genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or computational improvements [@Leonard2017GlowSoftware].

<!--maybe add "Julia's", "R's", "Python's" before the software names?-->
Software to model connectivity based on resistance surfaces include *ConScape.jl* [@vanMoorter2023AcceleratingAdvances], *Omniscape.jl* [@Landau2021OmniscapeJl], *lconnect* [@Mestre2023LconnectPackage], *grainscape* [@Chubaty2020ToolboxGrainscape], or *Circuitscape* [@Anantharaman2019CircuitscapeJulia].

### 2024-09-13 Sections/05-simulations.qmd
adds some comments, changes, etc

Pattern-based approaches simulate landscapes without assuming any underlying abiotic or biotic processes [i.e., neutral landscape model\; @Gardner1987NeutralModels].
The earliest neutral models are based on percolation theory and assign LULC types randomly to cells in the landscape[@Gardner1987NeutralModels] or hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992HierarchicalNeutral].
Landscapes characterized by continuous values can be simulated by fractal models, such as Brownian motion [@Palmer1992CoexistenceSpecies].
Borrowing from computer graphics, more recent neutral landscape models make use of spectral synthesis [e.g., Perlin noise\; @Etherington2022PerlinNoise] or binary space partitioning [@Etherington2022BinarySpace].
Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities using based on least-cost networks [@Etherington2024LeastcostNetwork].
In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral] or are based on target values optimization [@vanStrien2016ImprovedNeutral; @Justeau-Allaire2022FragmentedLandscape].
In order to make multiple neutral landscape model algorithms accessible, several software libraries are available in prominent programming languages including Python [NLMpy\; @Etherington2015NLMpyPython], R [NLMR; @Sciaini2018NLMRLandscapetools], Julia [NeutralLandscapes.jl; @Poisot2023NeutralLandscapesJulia], or stand-alone software [GradientLand; @Cambui2015GradientLandSoftware]

Contrastingly, processes-based approaches explicitly include abiotic or biotic pattern-forming processes [i.e., landscape generators\; @Langhammer2019AgriculturalLandscape; @Salecker2019EFForTSLGrafLandscape].
Several earlier landscape generators are based on cellular automata models and are able to simulate urban growth [@Clarke1998LoosecouplingCellular] or deforestation [@Soares-Filho2002DinamicaStochastic].
More recent landscape generators allow to simulate patchy landscapes based on Gibbs processes [@Gaucherel2006NeutralModels], deforestation based on road and agricultural fields access [@Peer2013SimpleProcessbased], agricultural areas dominated by smallholders [@Salecker2019EFForTSLGrafLandscape], or vegetation surround watering points in semi-arid savanna rangelands [combining pattern- and process-bases approaches\; @Hess2020PioLaGPiosphere].

@@

## Digital twins

Digital Twins include three connected elements, namely a physical object or system, its digital representation, and a data exchange between the former two [@Jones2020CharacterisingDigital; @deKoning2023DigitalTwins] and there is an increasing interest and use of them in both the industry and academia [@Jones2020CharacterisingDigital].
The need for timely and evidence-based decision making in combination with increasing data availability makes them a powerful tool for ecology [@deKoning2023DigitalTwins].
Digital twins are increasingly applied in agricultural landscapes research, including topics such as crop and land cultivation, livestock farming, controlled environment farming, or fertilization management [@Cesco2023SmartAgriculture; @Purcell2023DigitalTwins].
Further applications include planning of rural ecological landscapes [@Tan2024DigitalTwin], or exploration of relationships between urban expansion and vegetation coverage [@Zhao2022ApplyingDigital].
The use of general data processing, statistical and mechanistic models, or artificial intelligence allows them to constantly update the digital representation and analyze, visualize, or predict the physical counterpart.
Thus, Digital Twins can provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or identify uncertainties and information gaps [@deKoning2023DigitalTwins].

>>>>

Pattern-based approaches simulate landscapes without assuming any underlying abiotic or biotic processes [i.e., neutral landscape model\; @Gardner1987NeutralModels].
The earliest neutral models are based on percolation theory and assign LULC types randomly to cells in the landscape [@Gardner1987NeutralModels] or hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992HierarchicalNeutral].
Landscapes characterized by continuous values can be simulated by fractal models, such as Brownian motion [@Palmer1992CoexistenceSpecies].
Borrowing from computer graphics, more recent neutral landscape models make use of spectral synthesis [e.g., Perlin noise\; @Etherington2022PerlinNoise] or binary space partitioning [@Etherington2022BinarySpace].
Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities using based on least-cost networks [@Etherington2024LeastcostNetwork].
In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral] or are based on target values optimization [@vanStrien2016ImprovedNeutral; @Justeau-Allaire2022FragmentedLandscape].
In order to make multiple neutral landscape model algorithms accessible, several software libraries are available in prominent programming languages including Python (NLMpy\; @Etherington2015NLMpyPython), R (NLMR; @Sciaini2018NLMRLandscapetools), Julia (NeutralLandscapes.jl; @Poisot2023NeutralLandscapesJulia), or stand-alone software (GradientLand; @Cambui2015GradientLandSoftware).

Contrastingly, processes-based approaches explicitly include abiotic or biotic pattern-forming processes [i.e., landscape generators\; @Langhammer2019AgriculturalLandscape; @Salecker2019EFForTSLGrafLandscape].
Several earlier landscape generators are based on cellular automata models and are able to simulate urban growth [@Clarke1998LoosecouplingCellular] or deforestation [@Soares-Filho2002DinamicaStochastic].
More recent landscape generators allow to simulate patchy landscapes based on Gibbs processes [@Gaucherel2006NeutralModels], deforestation based on road and agricultural fields access [@Peer2013SimpleProcessbased], agricultural areas dominated by smallholders [@Salecker2019EFForTSLGrafLandscape], or vegetation surround watering points in semi-arid savanna rangelands [combining pattern- and process-bases approaches\; @Hess2020PioLaGPiosphere].
<!--JN: software??-->

@@

## Digital Twins

The Digital Twins concept include three connected elements, namely a physical object or system, its digital representation, and a data exchange between the former two [@Jones2020CharacterisingDigital; @deKoning2023DigitalTwins] and there is an increasing interest and use of them in both the industry and academia [@Jones2020CharacterisingDigital].
The need for timely and evidence-based decision making in combination with increasing data availability makes them a powerful tool for ecology [@deKoning2023DigitalTwins].
Digital twins are increasingly applied in agricultural landscapes research, including topics such as crop and land cultivation, livestock farming, controlled environment farming, or fertilization management [@Cesco2023SmartAgriculture; @Purcell2023DigitalTwins].
Further applications include planning of rural ecological landscapes [@Tan2024DigitalTwin], or exploration of relationships between urban expansion and vegetation coverage [@Zhao2022ApplyingDigital].
The use of general data processing, statistical and mechanistic models, or artificial intelligence allows them to constantly update the digital representation and analyze, visualize, or predict the physical counterpart.
<!-- JN: "Digital Twins can" or "Digital Twins will" or "Digital Twins aims to"? -->
Thus, Digital Twins can provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or identify uncertainties and information gaps [@deKoning2023DigitalTwins].
<!--JN: software??-->

### 2024-09-13 Sections/07-genetics.qmd
adds some comments, changes, etc

Advances in molecular methodologies and  synergistic developments of bioinformatic and computational approaches to analyze large-scale genomic data have recently enabled the integration of genome-wide data with spatial ecological data. <!-- MH: Ref missing -->.
This shift towards landscape genomics [@Cushman2018EditorialLeast; @Storfer2018NavigatingInterface] comes with novel utilities through which researchers can apply genomic data across spatio-temporal scales, including the integration of historic or ancient DNA [@Fenderson2020SpatiotemporalLandscape], or simulated present-day and future population genomic data [@House2018EvaluatingMethods; @Chevy2024PopulationGenetics]. 
Analytical frameworks such as the *FOLDS* model [gene flow, genetic offsets, genetic load, dispersal and SDMs, @Aguirre-Liguori2021EvolutionaryGenomics], constrained coordination where covarying sets of genotypes are correlated with multivariate environments through redundancy analysis [RDA, @Capblancq2021RedundancyAnalysis], and spatial simulations provide promising potential for future landscape genomic studies. 
Such studies may explore interactions between complex evolutionary processes, including demography, multidimensional (e.g., strength and direction) gene flow and migration across spatio-temporal scales, genetic differentiation across landscapes, genomic load and adaptive potential, and the interpretation of these evolutionary processes in context of landscape ecology [@Balkenhol2017LandscapeGenomics; @House2018EvaluatingMethods; @Fenderson2020SpatiotemporalLandscape; @Marcus2021FastFlexible; @Chevy2024PopulationGenetics].

>>>>

Advances in molecular methodologies and  synergistic developments of bioinformatic and computational approaches to analyze large-scale genomic data have recently enabled the integration of genome-wide data with spatial ecological data.<!-- MH: Ref missing -->.
This shift towards landscape genomics [@Cushman2018EditorialLeast; @Storfer2018NavigatingInterface] comes with novel utilities through which researchers can apply genomic data across spatio-temporal scales, including the integration of historic or ancient DNA [@Fenderson2020SpatiotemporalLandscape], or simulated present-day and future population genomic data [@House2018EvaluatingMethods; @Chevy2024PopulationGenetics]. 
Analytical frameworks such as the *FOLDS* model [gene flow, genetic offsets, genetic load, dispersal and SDMs, @Aguirre-Liguori2021EvolutionaryGenomics], constrained coordination where covarying sets of genotypes are correlated with multivariate environments through redundancy analysis [RDA, @Capblancq2021RedundancyAnalysis], and spatial simulations provide promising potential for future landscape genomic studies. 
Such studies may explore interactions between complex evolutionary processes, including demography, multidimensional (e.g., strength and direction) gene flow and migration across spatio-temporal scales, genetic differentiation across landscapes, genomic load and adaptive potential, and the interpretation of these evolutionary processes in context of landscape ecology [@Balkenhol2017LandscapeGenomics; @House2018EvaluatingMethods; @Fenderson2020SpatiotemporalLandscape; @Marcus2021FastFlexible; @Chevy2024PopulationGenetics].
<!--JN: software??-->
